---
title: "M4: Outliers and Missing Data - Wind Speed Data"
author: "Luana Lima"
output: pdf_document
---


## Setting R code chunk options

First R code chunk is used for setting the options for all R code chunks. The choice echo=TRUE means both code and output will appear on report, include = FALSE neither code nor output is printed.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages and initializing

Second R code chunk is for loading packages. By setting message = FALSE, the code will appear but not the output. 

```{r package, message=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)
library(Kendall)
library(tseries)

#New packages for M4
#install.packages("outliers")
library(outliers)
#install.packages("tidyverse")
library(tidyverse)

```

## Importing and preparing data

Today we will work with wind speed data. 

```{r}

#Importing time series data from text file#
wind_data <- read.csv(file="/home/guest/TSA_Sp26/Data/Wind_Speed_PortArthurTX_Processed.csv",header=TRUE,stringsAsFactors = TRUE)

#creating data object
wind_data$DATE <- ym(wind_data$DATE)

```


## Transforming data into time series object

Many of the functions we will use require a time series object. You can transform your data in a time series using the function *ts()*. 

```{r}
ts_wind_data <- ts(wind_data$AWND.1,frequency=12) 

autoplot(ts_wind_data) +
  ylab("Average Wind Speed") 
```


## Using pre-built function to clean series

tsclean() from package 'forecast' identify and replace outliers and missing values in a time series using linear interpolation. This should be used with some caution!!

```{r}

clean_wind_data <- tsclean(ts_wind_data)  #object should be a time series - uses local mean - dont need to worry about trend and seasonality - remove outlier funct you would need to deseason and detrend 

autoplot(clean_wind_data, series="Wind Clean") +
  autolayer(ts_wind_data, series="Wind Raw") +
  ylab("Average Wind Speed") 

#Note we only replaced missing data, algorithm did not identify any outliers for this data set.
  

```

If you want to handle missing data, but do not want to remove outliers, here are two other options that will replace them using linear interpolation.

`forecast::na.interp()`
`zoo::na.approx()`


```{r}
summary(wind_data$AWND, frequency = 12, start()) #wil show how many NAs are in what column 
    #summary(wind_data$AWND, frequency = 12, start())

#how to identify if they removed a row data point just bc there was NA in it 
#if it mises in july or auguest and not until dec you know there are missing values 
tail(ts_wind_data) #when is last observation andyoull know row missing 
ts_wind_data
```





